D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-05 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Some more small changes to the initial parts of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In broad terms,
OOTA occurs when a group of threads load from each others' stores
and each thread's store depends on the value returned by that thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air'',
but this is not necessary.

@@

Semantic dependencies are only one type of link in an OOTA cycle.
The other type extends from a given store to the load that returns the
value stored.
It is tempting to make the intuitive argument that the store must precede
the load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% @@@ This citation credits the formulation to Ali Sezgin, third author
% @@@ of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}
This is a fine definition and is consistent with the words in the C++
standard, but it has an intra-thread problem exemplified by the following
code:
\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r3, memory_order_relaxed);
\end{verbatim}
\end{quote}
This is an elaboration of \co{thread2()} from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation is within its rights to
note that line~3 might well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if the source code
had instead been as follows:
\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   // int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r2, memory_order_relaxed);
\end{verbatim}
\end{quote}
Here line~3 has been optimized away in favor of line~4 storing the same value
in \co{r2} that was stored to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Since we no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of an OOTA cycle again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores
	$W_i$~($i = 0$, \ldots,~$m$),
	where $W_i$ semantically depends on loads
	$\{R_{i,j}\}$~($j = 0$, \ldots,~$n_i$),
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This makes OOTA more complicated than a simple cycle but
we will continue to refer to OOTA cycles out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

Consider the following \co{if} statement:
\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = r1;
else
    z = r1;
\end{verbatim}
\end{quote}
Here there is a semantic dependency from \co{x}, but in some executions
it extends to \co{y} and in others to \co{z}.
This is an example of the load affecting not the value of
a given store, but rather whether or not that store is executed at all.

@@

The C++ standard describes the execution of a program in terms of
``a parameterized nondeterministic abstract machine'' in 4.1.2p1
(\co{intro.abstract}).
This description specifies how the abstract machine carries out the
operations of a source program in great, but not complete, detail:
\begin{itemize}
\item	Some of the abstract machine's characteristics are
	implementation defined, including things like the number of
	bits in the various integer types
	or whether the \co{char} type is signed.
\item	Some aspects of an execution are unspecified or nondeterministic,
	including things like the order of evaluation of the operands
	of most binary operators or of the arguments in a function call.
	Implementations may choose from a set of allowed behaviors.
\item	Some actions are deemed to have undefined results; the standard
	says essentially nothing about programs that can give rise to
	undefined behavior.
\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
\item	Input and output are not described in any detail.
\end{itemize}
In addition to these points, the standard does not specify which store
an atomic load must read from, beyond requiring that the overall
pattern of loads and stores be consistent with the C++ memory model.
We assume that programs will not indulge in any other computations that
could be expected to vary spontaneously from one execution to another,
such as basing a dependency on the time of day or a process ID.

@@

The same is theoretically true for the nondeterministic aspects of an
execution.
Consider this example, with \co{i} initially zero:
\begin{quote}
\begin{verbatim}
r1 = ++i / ++i;
x = r1 * z;
\end{verbatim}
\end{quote}
Because the order of evaluation of the division's arguments is undetermined,
the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
($2/1$).
In the former case there is no semantic dependency from \co{z} to \co{x},
but in the latter case there is.\footnote{
	Thanks to Peter Sewell for pointing out this possibility.}
(According to the current version of the standard, conflicting side effects
in unsequenced expressions constitute undefined behavior.
However there are proposals to make them defined in both
C++ and C,\footnote{
	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
which would render this theoretical possibility quite real.)

The abstract executions we use will be fully specified.
This means that all the missing information must be supplied:
the implementation-defined characteristics, the selections for the
nondeterministic pathways, and most notably, for each load, the store
from which it reads and the value of the load.
We ignore issues of signal handlers and I/O;
in any case our litmus-test programs don't use them.
The totality of this information---along with the program's source
code, of course---determines within each thread a unique, linearly
ordered series of steps to be carried out by the abstract machine.
(However, there is no ordering relation between steps carried out
in different threads.  Even if a load in one thread reads from a
store in another thread, we don't require the store to come before
the load.)

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).\footnote{
	This requirement is the standard's ``as-if'' rule.}
This means:\begin{itemize}
\item	The program's output must be the same as that of the abstract
	execution.
\item	Volatile accesses ``are evaluated strictly according to the
	rules of the abstract machine'' [4.1.2p6.1 (\co{intro.abstract})].
\item	(There is a condition on the interleaving of input and output,
	which does not matter for our purposes.)
\end{itemize}
We will say that the abstract execution is \emph{realized by} the
hardware execution.

@@

Exactly what the standard's restriction on volatile accesses means
isn't entirely clear.
The handling of volatiles, as understood by compiler developers, has
been described as more folklore or a gentlemen's agreement than
anything else.
We will take it to imply that a compiler may not
invent, omit, or merge volatile accesses, in a sense to be discussed below.

>>>>

In broad terms,
OOTA occurs theoretically when a group of threads load from each others' stores
and each thread's store depends on the value returned by that thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air'',
but this is not necessary.

@@

Semantic dependencies are only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to make the intuitive argument that the store must precede
the load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% @@@ This citation credits the formulation to Ali Sezgin, third author
% @@@ of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}

This is a fine definition and is consistent with the words in the C++
standard, but it has an intra-thread problem exemplified by the following
code:
\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r3, memory_order_relaxed);
\end{verbatim}
\end{quote}
This is an elaboration of \co{thread2()} from
Listing~\ref{lst:Simple OOTA}
that adds \co{z} along with lines~3 and~4.
The problem is that a C++ implementation is within its rights to
note that line~3 might well execute immediately after line~2, giving
other threads no chance to modify \co{y} in between.
Such an implementation might therefore behave as if the source code
had instead been as follows:
\begin{quote}
\scriptsize
\begin{verbatim}
 1   int r2 = atomic_load_explicit(&x, memory_order_relaxed);
 2   atomic_store_explicit(&y, r2, memory_order_relaxed);
 3   // int r3 = atomic_load_explicit(&y, memory_order_relaxed);
 4   atomic_store_explicit(&z, r2, memory_order_relaxed);
\end{verbatim}
\end{quote}
Here line~3 has been optimized away in favor of line~4 storing the same value
to \co{z} that was stored to \co{y} by line~2.
And given that the load from \co{y} no longer exists, it cannot possibly
act as a temporal constraint.

@@

Since we can no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of an OOTA cycle again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores
	$W_i$~($i = 0$, \ldots,~$m$),
	where $W_i$ semantically depends on loads
	$\{R_{i,j}\}$~($j = 0$, \ldots,~$n_i$),
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This makes OOTA more complicated than a simple cycle but
we will continue to refer to OOTA cycles out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

Consider the following \co{if} statement:
\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = r1;
else
    z = r1;
\end{verbatim}
\end{quote}
Here there is a semantic dependency from \co{x}, but in some executions
it extends to \co{y} and in others to \co{z}.
This is an example of a load affecting not the value of
a given store, but rather whether or not that store is executed at all.

@@

The C++ standard describes the execution of a program in terms of
``a parameterized nondeterministic abstract machine'' in 4.1.2p1
(\co{[intro.abstract]}).
This description specifies how the abstract machine carries out the
operations of a source program in great, but not complete, detail:
\begin{itemize}
\item	Some of the abstract machine's characteristics are
	implementation defined, including things like the number of
	bits in the various integer types
	or whether the \co{char} type is signed.
\item	Some aspects of an execution are unspecified or nondeterministic,
	including things like the order of evaluation of the operands
	of most binary operators or of the arguments in a function call.
	Implementations may choose from a set of allowed behaviors.
\item	Some actions are deemed to have undefined results; the standard
	says essentially nothing about programs that can give rise to
	undefined behavior.
\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
\item	Input and output are not described in any detail.
\end{itemize}
In addition to these points, the standard does not specify which store
an atomic load must read from, beyond requiring that the overall
pattern of loads and stores be consistent with the C++ memory model.
We assume that programs will not indulge in any other computations that
could be expected to vary spontaneously from one execution to another,
such as basing a dependency on the time of day or a process ID.

@@

The same is true for the nondeterministic aspects of an execution.
Consider this example, with \co{i} initially zero:
\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

r1 = foo(++i, ++i);
x = r1 * z;
\end{verbatim}
\end{quote}
Because early C~implementers could not come to agreement, the standard
does not specify the order of evaluation of function arguments, so
the value calculated for \co{r1} might be zero ($1/2$ truncated) or two
($2/1$).
In the former case there is no semantic dependency from \co{z} to \co{x},
but in the latter case there is.\footnote{
	Thanks to Peter Sewell for pointing out this possibility.}

(According to the current version of the standard, conflicting side effects
in unsequenced subexpressions constitute undefined behavior,
although there are proposals to make them defined in both C++ and C.\footnote{
	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
Nevertheless, the example above is allowed because the order of evaluation
of arguments to a function call is ``indeterminately sequenced''
(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)

\medskip

The abstract executions we use will be fully specified.
This means that all the missing information must be supplied:
the implementation-defined characteristics, the selections for the
nondeterministic pathways, and most notably, for each load, the store
from which it reads and the value of the load.
We ignore issues of signal handlers and I/O;
in any case our litmus-test programs don't use them
(but see the discussion of volatile loads in
section~\ref{sec:Relation Between Abstract and Hardware Executions} below).
The totality of this information---along with the program's source
code, of course---determines within each thread a unique, linearly
ordered series of steps to be carried out by the abstract machine.
However, there is no ordering relation between steps carried out
in different threads.  Even if a load in one thread reads from a
store in another thread, we don't require the store to come before
the load.

@@

The C++ standard requires that for any valid implementation, when a
program runs its observable behavior must be the same as that of some
abstract execution of the source code given the same input (in the
absence of any abstract executions containing undefined behavior).\footnote{
	This requirement is the standard's ``as-if'' rule.}
This means:
\begin{itemize}
\item	The program's output must be the same as that of the abstract
	execution.
\item	Volatile accesses ``are evaluated strictly according to the
	rules of the abstract machine'' (4.1.2p6.1 \co{[intro.abstract]}).
\item	(There is a condition on the interleaving of input and output,
	which does not matter for our purposes.)
\end{itemize}
We will say that the abstract execution is \emph{realized by} the
hardware execution.

@@

Exactly what the standard's restriction on volatile accesses means
isn't entirely clear.
The handling of volatiles, as understood by compiler developers, has
been described as more folklore or a gentlemen's agreement than
anything else.
To help guide C++ users and implementers, the standard adds these
suggestive comments (9.2.9.2p5 and~6 \co{[dcl.type.cv]}):
\begin{quote}
	The semantics of an access through a volatile glvalue are
	implementation-defined.

	\co{volatile} is a hint to the implementation to avoid aggressive
	optimization involving the object because the value of the object
	might be changed by means undetectable by an implementation.
\end{quote}
Following the folklore, we interpret these comments to mean that
volatile accesses should be treated as a form of I/O,
implying that volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which must return the value of the store they read from).
As a corollary, the rfe relation does not apply to volatile loads
and hence the accesses in an OOTA cycle must not be volatile.
We also understand that a compiler may not
invent, omit, merge, or reorder volatile accesses,
in a sense to be discussed below.
